Novelty Detection Using Graphical Models for Semantic Room Classification

### 2011-05-16 semantic-nd/article.tex
Added images and some texts fixed to make it more coeherent.
Eg.: call inputs as sensed features all the time.

A \emph{factor graph} $G$ is a bipartite graph connecting a set of nodes $V_G$ representing
random variables and factors $\phi_G$.
Each factor represents a function dependent only on the variables to where its connected.
As so given factor graph $G$ can be seen as description of a distribution function over a set
of variables obtained by the product of all the factors. In order to represent a probability
a normalization factor $Z$ needs to be introduced:

\begin{equation}
P_G(X_V) = \frac{1}{Z}\prod_{}{\phi(...)},\qquad Z = \sum\prod{\phi(...)}
\end{equation}

Describing the distribution function in terms of graphs allows to use algorithms such as the
Sum-Product to efficiently calculate marginals on any given subset $x$ of variables by
exploiting conditional independency between variables.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Novelty Detection}
% Introduce novelty detection
Novelty detection deals with detecting that a new input was generated by a class
other than the ones the system knows about.
It is harder than classification as only positive samples of a class are available
rendering normal classification methods invalid.

@@

\subfloat[Graph structure $G$ used to model conditional probability graph.]{... conditional probability}
\qquad
\subfloat[Graph structure $U$ used to model unconditional probability graph.]{... unconditional probability}
\qquad
\subfloat[Graph structure $I$ used to model unconditional probability graph.]{... unconditional probability}

@@

\subfloat[3 sensed variables]{\includegraphics[width=0.40\textwidth]{results/synthetic-3features.pdf}}
\qquad
\subfloat[5 sensed variables]{\includegraphics[width=0.40\textwidth]{results/synthetic-5features.pdf}}

\subfloat[10 sensed variables]{\includegraphics[width=0.40\textwidth]{results/synthetic-10features.pdf}}
\qquad
\subfloat[50 sensed variables]{\includegraphics[width=0.40\textwidth]{results/synthetic-50features.pdf}}

\caption{\label{fig:synthetic-roc-breakdown}ROC curves plotted showing performance of the
         presented novelty detection method on graphs generated for different amount of
         sensed variables.}
\end{figure}

@@

The performance of the independent threshold decreases as the number of sensed
variables increases. This is easily explained by the fact that the graph $I$ is not
able to model the existent dependence between the variables and that becomes obvious
as the number of variables increases (Eg.: graph $I$ perfectly models $P(x)$ in the
case where only 1 variable is present).

The uniform threshold shows a very poor performance specially on small size samples
where its performs almost no better than random.
It performance increases as the size of sensed variables increases but nonetheless
its always very small when compared to how optimal a threshold could be.
This shows that at least under semantic distributions as the one modelled with the
synthetic data assuming an uniform unconditional distribution is a very bad choice.

>>>>

A \emph{factor graph} is a bipartite graph connecting a set of nodes representing
random variables and factors $\phi_G$.
Each factor represents a function dependent only on the variables to where its connected.
As so given factor graph can be seen as description of a distribution function over a set
of variables obtained by the product of all the factors. In order to represent a probability
a normalization factor needs to be introduced:

\begin{equation}
P_G(x) = \frac{1}{Z}\prod_{}{\phi(...)},\qquad Z = \sum\prod{\phi(...)}
\end{equation}

Describing the distribution function in terms of graphs allows to use algorithms
such as the Sum-Product~\citep{kschischang2001factor} to efficiently calculate
marginals on any given subset $x$ of variables by exploiting conditional
independence between variables.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Novelty Detection}
% Introduce novelty detection
Novelty detection deals with detecting that a new input was generated by a class
other than the ones the system knows about \citep{markou2003novelty}.
It is harder than classification as only positive samples of a class are available
rendering normal classification methods invalid.

@@

\subfloat[Graph structure $G$.]{
         \includegraphics[width=0.27\textwidth]{figures/simple-cond-graph.pdf}}
\qquad
\subfloat[Graph structure $U$.]{
         \includegraphics[width=0.27\textwidth]{figures/simple-uniform-graph.pdf}}
\qquad
\subfloat[Graph structure $I$.]{
         \includegraphics[width=0.27\textwidth]{figures/simple-independent-graph.pdf}}

@@

\subfloat[3 sensed features]{\includegraphics[width=0.40\textwidth]{results/synthetic-3features.pdf}}
\qquad
\subfloat[5 sensed features]{\includegraphics[width=0.40\textwidth]{results/synthetic-5features.pdf}}

\subfloat[10 sensed features]{\includegraphics[width=0.40\textwidth]{results/synthetic-10features.pdf}}
\qquad
\subfloat[50 sensed features]{\includegraphics[width=0.40\textwidth]{results/synthetic-50features.pdf}}

\caption{\label{fig:synthetic-roc-breakdown}ROC curves plotted showing performance of the
         presented novelty detection method on graphs generated for different amount of
         sensed features.}
\end{figure}

@@

The performance of the independent threshold decreases as the number of sensed
features increases. This is easily explained by the fact that the graph $I$ is not
able to model the existent dependence between the features and that becomes obvious
as the number of features increases (Eg.: graph $I$ perfectly models $P(x)$ in the
case where only 1 feature is sensed).

The uniform threshold shows a very poor performance specially on small size samples
where its performs almost no better than random.
It performance increases as the size of sensed features increases but nonetheless
its always very small when compared to how optimal a threshold could be.
This shows that at least under semantic distributions as the one modelled with the
synthetic data assuming an uniform unconditional distribution is a very bad choice.
